### 2018-03-15 paper/3-abstractions.tex
Work on S4

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: (Eq k, Hashable v) => Build Monad () k v
busy task key store = execState (compute key) store
  where
    compute k = case task compute k of
        Nothing  -> do { s <- get; return (getValue s k) }
        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
\end{minted}
\vspace{1mm}

>>>>

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: (Eq k, Hashable v) => Build Monad () k v -- Drop Hashable? It's confusing here
busy task key store = execState (compute key) store
  where
    compute k = case task compute k of
        Nothing  -> do { s <- get; return (getValue s k) }
        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
\end{minted}
\vspace{1mm}

### 2018-03-15 paper/4-build.tex
Work on S4

\section{Build systems, abstractly}\label{sec-build}

In this section we describe our \hs{Build} abstraction, defined as:

\begin{minted}{haskell}
type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
\end{minted}

The build function takes 4 type arguments: \hs{c} is the constraint used for \hs{Compute} (typically either \hs{Applicative} or \hs{Monad}), \hs{i} is the information stored by the build system and \hs{k}/\hs{v} are the key/value types as described in \S\ref{sec-vocabulary}. Given a build system \hs{build}:

\begin{minted}{haskell}
build compute key info store = (info', store')
    where ...
\end{minted}

The function \hs{build} is responsible for ensuring that \hs{key} and its dependencies are up-to-date in \hs{store'} starting from \hs{store}. As part of that work, the \hs{build} function \textit{may} store other information in \hs{info'}, which will be provided as \hs{info} in future calls to \hs{build}. The \hs{info} parameter will be \hs{Nothing} on the very first call to \hs{build}. We assume that \hs{store} is total, where "missing files" can be represented by some sentinel value of \hs{v} if necessary.

All the \hs{Build} types we define are polymorphic in both \hs{k} and \hs{v} (perhaps adding constraints to them), and have specific instantiations of \hs{c} and \hs{i}. How to implement \hs{Build} can be captured by the idea:

\begin{quote}
A build system must build \underline{out-of-date keys} while \underline{respecting dependency orderings}.
\end{quote}

We have underlined two different phrases in this sentence, and tackle each aspect separately.

\subsection{Respecting dependency orderings}
\label{sec-dependency-orderings}

If the \hs{Compute} function for \hs{k} calls the functional argument of \hs{d} then the build system is responsible for ensuring that \hs{d} is built \textit{before} \hs{k}. Looking at the build systems in \S\ref{sec-background} we have observed 3 distinct approaches to respecting dependency orderings. In this section of the paper we'll explore what properties are a consequence and how they operate.

\subsubsection{Topological sort}

The topological approach creates a linear ordering, which when followed, ensures each build is correct. Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort. However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{sec-deps}.

@@

The topological sort approach has two downside - it is limited to \hs{Applicative} build systems and requires a fresh topological sort each time - while the actions themselves may be incremental the pre-processing is not. An alternative approach is to keep the topological order between runs and assume it to be correct, but if execution determines it is wrong, reorder.

This requires a way to abort computations that have failed. It's also not minimal in the sense that computations start, may do some meaningful work, then abort. However, in the case of an \hs{Applicative} system, that work is zero. It does require the ability to terminate a running computation.

@@

An alternative way is to simply build dependencies when they are requested. By combining that with a transient store of which keys have been built in this execution, you can describe dependencies in a simple way. You need a way to suspend a running action, which can be done with cheap green-threads and blocking (the original approach of Shake) or with a continuation-passing style (what Shake does now). An alternative approach to suspending a computation is to abort it and restart it again later, at the cost of doing additional work.

\subsection{Determining what is out-of-date}

The second part, determining what to rebuild, can be addressed in one of three fundamental ways, with a number of tweaks and variations within them.

\subsubsection{A dirty bit}
\label{sec-dirty-bit}

The idea of a dirty bit is to have one piece of information, whether a key is dirty or clean. After a run, all bits are set to clean. When a run starts anything that changed between the two states is marked dirty - by marking additional things dirty/clean the build system can track what needs to rebuild. When reaching a key, if it and all its dependencies are clean, the key does not need recomputing.

\Excel models the dirty bit approach most directly, having an actual dirty bit associated with each cell. Since Excel is the only manner by which a cell may be changed by the user, it can dirty the bits as necessary. When rebuilding if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and marked dirty itself. The only wrinkle in this scheme is that \Excel supports monadic build systems, and does not separately record what rebuilt, so it has to approximate in this case.

\Make uses file modification times, and compares files to their dependencies, which can be thought of as a dirty bit which is set when a file is newer than its dependencies. The interesting property of this dirty bit is that it is not under control of \Make, but an existing bit that is repurposed. In particular, updating a file automatically clears its dirty bit, and automatically sets the dirty bit of the nodes depending on it. One thing \Make does require is that file timestamps only go forward in time - something that can be violated by some backup software.

With a dirty bit it is possible to achieve minimality. To achieve cut-off it would be important to not set the dirty bit after a computation that did not change the value. \Excel could model this approach, but does not. In contrast, \Make cannot model cut-off - to do so it would have to mark the node clean (so it didn't rebuild in the next run) and at the same time not mark the things it depends on dirty - an impossible task with only the ability to update to the latest modification time.

>>>>

\section{Build Systems \`a la Carte}\label{sec-build}

The focus of this paper is on a variety of implementations of \hs{Build}, given
a client-supplied implementation of \hs{Task}. That is, we are going to take
\hs{Task} as given from now on, and explore \hs{Build} variants: first
abstractly (in this section) and then concretely in~\S\ref{sec-implementations}.

Let \hs{result}~\hs{=}~\hs{build}~\hs{task}~\hs{key}~\hs{store}. The build
system \hs{build} is responsible for ensuring that \hs{key} and its dependencies
are up-to-date in \hs{result} starting from \hs{store} (see the formal definition
of correctness in~\S\ref{sec-build-correctness}). As part of that work,
\hs{build} \emph{may} store auxiliary information \hs{info} of type \hs{i} in
the \hs{store}, which will be provided to future calls of \hs{build}.

All the build systems we define are polymorphic in both \hs{k} and \hs{v}
(perhaps adding constraints to them), and have specific instantiations of \hs{c}
and \hs{i}. As an example, recall that the \hs{busy} build system defined
in~\S\ref{sec-general-build}, has \hs{c}~\hs{=}~\hs{Monad}, i.e. it supports
monadic tasks, and \hs{i = ()}, i.e. it does not store any build information.

According to the definition of minimality~\ref{def-minimal}, a minimal build
system must \textbf{rebuild only out-of-date keys} and at most once. The only
way to achieve the ``at most once'' requirement is to \textbf{build all keys in
the order that respects their dependencies}.

We have emphasised two different phrases above, and tackle each aspect separately.

\subsection{Respecting the dependency order}
\label{sec-dependency-orderings}

The build systems overview (\S\ref{sec-background-summary}) highlighted three
distinct approaches to respecting the dependency order. This subsection explores
their properties and possible implementations.

\subsubsection{Topological}

The topological approach computes a linear order, which when followed, ensures
the \hs{build} is correct regardless of the initial \hs{store}. Given a function
from a key to its dependencies, and the output \hs{key}, you can compute the
linear order by first taking the transitive closure of \hs{key}'s dependencies,
and then computing its topological sort. However, as we have seen
in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task,
which requires the build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling
out dynamic dependencies.

@@

The topological approach has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time -- while the
actions themselves may be incremental the pre-processing is not. An alternative
approach is to store the topological order between build runs and assume it to
be correct, but if the build determines it is wrong, reorder.

This requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that tasks start, may do some
meaningful work, then abort. However, in the case of an \hs{Applicative} system,
that work is zero.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) is to simply build dependencies when they are
requested. By combining that with a transient store of which keys have already
been built, you can obtain a minimal build system.

This requires a way to suspend a running task, which can be done with cheap
green threads and blocking (the original approach of \Shake) or with a
continuation-passing style (what \Shake does now). An alternative approach to
suspending a task is to abort it and restart it again later, at the cost
of doing additional work.

\subsection{Determining out-of-date keys}

The second aspect, determining what to rebuild, can be addressed in one of three
fundamental ways, with a number of tweaks and variations within them.

\subsubsection{A dirty bit}\label{sec-dirty-bit}

The idea of a dirty bit is to have one piece of information, whether a key is
\emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
next build starts anything that changed between the two states is marked
dirty~--~by marking additional things dirty/clean the build system can track
what needs to rebuild. When reaching a key, if it and all its dependencies are
clean, the key does not need recomputing.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell. Since \Excel is the only manner by which a cell may
be changed by the user, it can mark cells dirty as necessary. When rebuilding,
if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and
marked dirty itself.

% AM: I didn't understand this bit, so commented it out.
% The only wrinkle in this scheme is that \Excel supports
% monadic tasks, and does not separately record the rebuilt, so it has
% to approximate in this case.

\Make uses file modification times, and compares files to their dependencies,
which can be thought of as a dirty bit which is set when a file is newer than
its dependencies. The interesting property of this dirty bit is that it is not
under control of \Make, but an existing bit that is repurposed. In particular,
rebuilding a file automatically clears its dirty bit, and automatically sets
the dirty bit of the nodes depending on it. One thing \Make does require is that
file timestamps only go forward in time -- something that can be violated by
backup software.

With a dirty bit it is possible to achieve minimality. To achieve early cutoff
(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
after a computation that did not change the value. \Excel could model this
approach, but does not. In contrast, \Make cannot model early cutoff -- to do so
it would have to mark the node clean (so it didn't rebuild in the next run) and
at the same time not mark the things it depends on dirty -- an impossible task
with only the ability to update to the latest modification time.
